---
title: "Illustrating Instability in Synthetic Data"
author: "Biyonka Liang, Arun Ramamurthy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Illustrating Instability in Synthetic Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(stablr); library(glue); library(dplyr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(108)
```

# Overview of Stability

What is stability, Biyonka?

# Illustrative Example I: Low-dimensional Collinearity

```{r}
## Generate a dataframe of 2 uniform covariates and a response, y = x_1 + x_2 + N(0, 1)
stable_data <- 
  stablr::generate_data(n = 100, k = 2, 
                        covariate_generation_fn = purrr::partial(stablr::generate_uniform_vector,
                                                                 bound = 10),
                        y_formula = x_1 + x_2,
                        y_noise_fn = purrr::partial(stablr::generate_normal_vector,
                                                    sd = 1))

## Generate a similar dataframe, with the same x_1, but a collinear x_2 = 2*x_1 + 5 + N(0, 1) and similarly computed y to above
unstable_data <- 
  dplyr::select(stable_data, x_1)  %>%
  stablr::augment_collinearity(new_col_formula = 2*x_1 + 5, 
                               new_col_noise_fn = purrr::partial(stablr::generate_normal_vector,
                                                                 sd = 1)) %>%
  stablr::augment_y(y_formula = x_1 + x_2,
                    y_noise_fn = purrr::partial(stablr::generate_normal_vector,
                                                sd = 1))

## Calculate Estimation Stability as presented in Yu 2013
stable_ES <-
  stable_data %>% stablr::ES.lm(y ~ ., V = 500, multiperturb_fn = multiperturb_norm, x_1, x_2) %>% round(5)
unstable_ES <-
  unstable_data %>% stablr::ES.lm(y ~ ., V = 500, multiperturb_fn = multiperturb_norm, x_1, x_2) %>% round(5)

print(
  glue::glue("For the same model formulation, the stable data shows an Estimation Stability of {stable_ES},\n
             whereas the unstable data shows an ES of {unstable_ES}.\n
             In this simulation, collinear data yields a reduction by {round((stable_ES - unstable_ES)/stable_ES, 4)*100} percent in Estimation Stability. ")
)
```

# Illustrative Example II: High-dimensional Collinearity

Example of use:

```{r}
iris %>% stablr::ES.lm(lm_formula = Sepal.Length ~ Petal.Length + Petal.Width,
                       V = 100, 
                       multiperturb_fn = multiperturb_norm,
                       Petal.Length, Petal.Width)
```


# Erase Me Later

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```
