---
title: "Illustrating Instability in Synthetic Data"
author: "Biyonka Liang, Arun Ramamurthy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Illustrating Instability in Synthetic Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(stablr); library(glue); library(dplyr);library(purrr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
set.seed(108)
```

# Overview of Stability

What is stability, Biyonka?

# Illustrative Example I: Low-dimensional Collinearity

```{r ex1}
## Generate a dataframe of 2 uniform covariates 
### and a response, y = x_1 + x_2 + N(0, 1)
stable_data <- 
  generate_data(n = 100, k = 2, 
                covariate_generation_fn = partial(generate_uniform_vector, bound = 10),
                y_formula = x_1 + x_2,
                y_noise_fn = partial(generate_normal_vector, sd = 1))

## Generate a similar dataframe, with the same x_1, 
### but a collinear x_2 = 2*x_1 + 5 + N(0, 1) and 
### and similarly computed y to above
unstable_data <- 
  select(stable_data, x_1)  %>%
  augment_collinearity(new_col_formula = 2*x_1 + 5, 
                       new_col_noise_fn = partial(generate_normal_vector, sd = 1)) %>%
  augment_y(y_formula = x_1 + x_2,
            y_noise_fn = partial(generate_normal_vector, sd = 1))

## Calculate Estimation Stability as presented in Yu 2013
stable_ES <-
  stable_data %>% 
  ES.lm(y ~ ., V = 500, multiperturb_fn = multiperturb_norm, x_1, x_2) %>% 
  round(5)
unstable_ES <-
  unstable_data %>% 
  ES.lm(y ~ ., V = 500, multiperturb_fn = multiperturb_norm, x_1, x_2) %>% 
  round(5)
```
```{r ex1_out, echo=F}
print(glue("For the same model formulation, the stable dataset shows an Estimation Stability of {stable_ES},\n
whereas the unstable dataset shows an ES of {unstable_ES}.\n
In this simulation, collinear data yield a reduction of {round((stable_ES - unstable_ES)/stable_ES, 4)*100} percent in Estimation Stability. "))
```


# Illustrative Example II: High-dimensional Collinearity

Example of use:

```{r}
iris %>% ES.lm(lm_formula = Sepal.Length ~ Petal.Length + Petal.Width,
                       V = 100, 
                       multiperturb_fn = multiperturb_norm,
                       Petal.Length, Petal.Width)
```


# Erase Me Later

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```
