---
title: "Illustrating Instability in Synthetic Data"
author: "Biyonka Liang, Arun Ramamurthy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Illustrating Instability in Synthetic Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(stablr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview of Stability

What is stability, Biyonka?

# Illustrative Example I: Low-dimensional Collinearity

# Illustrative Example II: High-dimensional Collinearity

Example of use:

```{r}
iris %>% stablr::ES.lm(lm_formula = Sepal.Length ~ Petal.Length + Petal.Width,
                       V = 100, 
                       multiperturb_fn = multiperturb_norm,
                       Petal.Length, Petal.Width)
```


# Erase Me Later

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```
